\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Add cell lines text

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within an ARG, and the relationship
between the two is subtle and important.
Note that many authors use the term ``marginal''
% could list loads here, but not much point.
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
which we avoid due to the potential confusion with the
statistical concept of marginal distributions
(potentially leading to interpretation as an integrated tree over many positions).

@@

>>>>

\section*{ARGs and local trees}
% TODO mention Yamato here? I'm sure there were some interesting points
% about local trees in that paper
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within an ARG, and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][AND OTHERS]{griffiths1996ancestral,minichiello2006mapping}
which we avoid due to the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

@@

\section{Cell lineages and ARGs}
\label{cell-lineages-and-args}
As we have argued, an important benefit of the gARG representation is that it
can concatenate multiple events by encoding the outcome of their effects
on a single node. In the case of recombination events, this manifests itself as
a single node with three or more parents (e.g.\ Fig.~\ref{fig-simplification}C, node \noderef{e}).
Here we make the argument that the same principle underlies ARGs containing
coalescence points that involve three or more children. These ``polytomies'' are generated not only
by Wright-Fisher dynamics such as those in Fig.~\ref{fig-simplification}C (see node \noderef{k}), but also
by life histories that involve ``burst'' reproduction events, which are often modelled by a lambda
coalecent process [AND OTHERS]\citep{wakely2008coalescent}.

At the base of this concept is a subtle and important point about which genomes the nodes in a gARG represent. 
In principle, we can interpret nodes as representing genomes
at any point in the life cycle, for instance, \citet{hudson1983properties} views nodes as representing gametes,
whereas in Fig.~\ref{fig-arg-in-pedigree}A, we present them as genomes within a diploid individual.
In reality, however, recombination and replication of DNA occur at a cellular level,
and correspond to the processes of meiosis and mitosis respectively. A comprehensive gARG would therefore
trace the process of inheritance not between diploid individuals, but between each cell during the
life cycle of an organism. This is illustrated in Fig.~\ref{fig-cell-lines}, where we show a simplified
schematic of cellular inheritance in a diploid population over four partially overlapping generations.
In this figure, thick black lines trace an ARG that describes the ancestry of the lower arm of the
four sampled chromosomes. The potential ARG nodes are shown as green chromosomal regions, and the ``events''
underlying the ARG are shown as red and blue squares.

The unusual feature of this particular example is that three ARG lineages coalesce in one individual, $D_10$
(who has three children). We know that the underlying coalescence process must consist of two successive bifurcations
as the only known method of reproducing DNA is by (semi-conservative) duplication. However, if we take each node in the
ARG to represents the generalised genome of an individual, rather than the specific genome of a cell within an individual,
we can only represent this as a instantanous trifucation or polytomy.

There is biological significance to this cellular-level inheritance process. Were a mutation to occur
during the first cell division of $D_10$, the two gametes produced by the cells
in the left half of $D_10$ could share a mutation not present in the right hand gamete. In other words,
the relationship between three (or more) gametes that originate, without recombination, from the same parental
chromosomes will nevertheless be strictly bifurcating; two of the gametes have a closer relationship with the
third more distant. Of course, unless mutations occur between successive rounds of cell division, the between-cell
relationships will be impossible to infer from the data, with subsequent implications for
using the eARG representation to store the results of inference.

As with concatenating multiple recombination events, the gARG representation allows us to concatenate within-individual
coalescence events such that we are not obliged to represent precisely the underlying cellular process. In humans and
other multicellular diploids, it seems most intuitive to take the nodes in the gARG to represent two haploid genomes within
a newly fertilised diploid individual. If this individual has 3 or more children, and these children feature in the
eventual ancestry of the samples, the gARG approach outlined in this paper means that they can be legitimately represented
by a polytomy in the local trees.
\begin{figure}
\begin{center}
    \includegraphics[width=0.8\textwidth]{illustrations/cell-lines}
\end{center}
\caption{\label{fig-cell-lines}
Cellular inheritance of a single chromosome in a diploid population.
Individuals (blue) contain diploid cells (white circles enclosing a homologous pair of chromosomes).
For clarity, only two rounds of mitotic germ-line cell division are shown per individual, and
meiosis is not illustrated in detail.
Lines show prospective inheritance paths for all chromosomes. Solid lines show all possible
retrospective ancestry paths for four chosen chromosomes (indicated by square black ``sampling events'')
sampled from 3 diploid individuals ($D_1$, $D_2$, $D_3$) in the current generation.
Ancestral recombination events and coalescence events are shown as red and blue squares respectively.
The ARG paths for the lower arm of the sampled chromosomes (green) is highlighted as a
think solid line and involves a single recombination event and four coalescence events
(highlighted as deep red and blue squares within individuals $D_5$, $D_10$, and $D_13$).
ARG lineages also show gametic genomes, contained within shaded circles.
As in Fig.~\ref{fig-arg-in-pedigree}A, inherited regions within the sampled chromosome arm are
shaded by the number of descendant samples.
}
\end{figure}
